---
title: "dplyr and tidyr"
output: html_notebook
---

```{r}
install.packages("devtools")
devtools::install_github("rstudio/EDAWR")
library(EDAWR)

```


Storms dataset

```{r}
View(storms)
storms$storm
```

cases dataset

```{r}
View(cases)
cases$country
names(cases)[-1]
unlist(cases[1:3, 2:4]) 
# return numeric vector, first num start from first row of data
# second num start from 2nd column which is 1st column of data
```

```{r}
pollution$city[c(1,3,5)]

```



tidyr: A package that reshapes the layout of tables

```{r}
library(tidyr)
?gather
?spread
```

**gather**

Collapses multiple columns into two columns:
1. a key column that contains the former column names
2. a value column that contains the former column cells

```{r}
head(cases)
```

```{r}
gather(cases, "year", "n", 2:4)
```

**spread**
```{r}
storm_united <- storms %>% separate(date, c("y", "m", "d"))
storm_united

```


**seperate**
```{r}
storm_united <- storms %>% separate(date, c("year", "month", "day"))
storm_united

```



**unite**

```{r}
storm_sep <- unite(storm_united, "date", year, month, day, sep = "-" )
storm_sep

```





### dplyr

```{r}
library(nycflights13)
library(dplyr)
```

**select**
Extract existing variables

```{r}
select(storms, storm, pressure)
#              cols to extract, if don't want to extract, use -col_name
#                               if a range, use wind:date
# select functions:
# - Select everything but
# : Select range
# contains() Select columns whose name contains a character string
# ends_with() Select columns whose name ends with a string
# everything() Select every column
# matches() Select columns whose name matches a regular expression
# num_range() Select columns named x1, x2, x3, x4, x5
# one_of() Select columns whose names are in a group of names
# starts_with() Select columns whose name starts with a character string
```

**filter**

Extract existing observations

```{r}
#
storms %>% filter(wind >= 50)
# 2 conditions
storms %>% filter(wind >= 50, storm %in% c("Alberto", "Alex"))

```


**mutate**
Derive new variables from existing ones
```{r}
mutate(storms, ratio = pressure / wind, inverse = ratio^-1)

mutate(storms, cummin(wind))
```

Useful mutate functions
* All take a vector of values and return a vector of values
** Blue functions come in dplyr

pmin(), pmax() Element-wise min and max
cummin(), cummax() Cumulative min and max
cumsum(), cumprod() Cumulative sum and product
between() Are values between a and b?
cume_dist() Cumulative distribution of values
cumall(), cumany() Cumulative all and any
cummean() Cumulative mean
lead(), lag() Copy with values one position
ntile() Bin vector into n buckets
dense_rank(), min_rank(),
percent_rank(), row_number() Various ranking methods

"Window" functions
pmin(), pmax() Element-wise min and max
cummin(), cummax() Cumulative min and max
cumsum(), cumprod() Cumulative sum and product
between() Are values between a and b?
cume_dist() Cumulative distribution of values
cumall(), cumany() Cumulative all and any
cummean() Cumulative mean
lead(), lag() Copy with values one position
ntile() Bin vector into n buckets
dense_rank(), min_rank(),
percent_rank(), row_number()


**summarise**
Change the unit of analysis
```{r}
pollution %>% summarise(median = median(amount), variance = var(amount))
```

Useful summary functions
min(), max() Minimum and maximum values
mean() Mean value
median() Median value
sum() Sum of values
var, sd() Variance and standard deviation of a vector
first() First value in a vector
last() Last value in a vector
nth() Nth value in a vector
n() The number of values in a vector
n_distinct() The number of distinct values in a vector
* All take a vector of values and return a single value
** Blue functions come in dplyr


**arrange**

```{r}
arrange(storms, desc(wind))
arrange(storms, wind, date)
```

**group_by**

```{r}

```

**pipeline**
short cut to type %>%:
cmd + shift + M


```{r}
pollution %>% 
  group_by(city) %>% 
  summarise(mean = mean(amount), sum = sum(amount), n = n()) %>% 
  arrange(mean)
```


Other:
 ungroup()
 
 
```{r}
tb %>%
  select(country, year, sex) %>% 
  group_by(country, year) 
```


#### Join data
bind_cols(x,y)
bind_rows(x1,x2)
union(x, y)
intersect(x,y)
setdiff(y, z)
left_join(songs, artists, by = "name")
left_join(songs2, artists2, by = c("first", "last"))
inner_join(songs, artists, by = "name")
semi_join(songs, artists, by = "name")
