---
title: "Two Variables"
output: html_notebook
---

## Scatter Plot

### jitter and alpha

```{r}

library(ggplot2)
pf <- read.csv('pseudo_facebook.tsv', sep = '\t')

ggplot(pf, aes(x=age, y = friend_count)) +
  geom_point(alpha = 1/20) +
  xlim(13,90)

```

```{r}
ggplot(pf, aes(x=age, y = friend_count)) +
  geom_jitter(alpha = 1/20) +
  xlim(13,90)
```

```{r}
ggplot(pf, aes(x=age, y = friend_count)) +
  geom_point(alpha = 1/20) +
  coord_trans(y = "sqrt")
  
```

```{r}
ggplot(pf, aes(x=age, y = friend_count)) +
  geom_point(alpha = 1/20, position = position_jitter(h=0)) +  # remove negative friend_count
  coord_trans(y = "sqrt") +
  xlim(13, 90)
  
```
```{r}
ggplot(pf, aes(x=age, y = friendships_initiated)) +
  geom_point(alpha = 1/20, position = position_jitter(h=0)) +  # remove negative friend_count
  coord_trans(y = "sqrt") +
  xlim(13, 90)
  
```


### Plot mean friend count vs. age using a line graph.

```{r}

pf.fc_by_age <- pf %>% 
  group_by(age) %>% 
  summarise(mean = mean(friend_count), median = median(friend_count), n=n()) %>%
  arrange(age)

ggplot(pf.fc_by_age, aes(x = age, y = mean))+
  geom_line(color = "blue") + 
  scale_x_continuous(breaks = seq(0,120,10))

```


## Overlaying summaries

```{r}
ggplot(aes(x = age, y = friend_count), data = pf) +
  geom_point(alpha = 0.05, position = position_jitter(h = 0), color = 'orange') +
  # coord_trans(y = 'sqrt') +
  geom_line(stat = 'summary', fun.y = mean) + 
  geom_line(stat = 'summary', fun.y = quantile, fun.args = list(probs = 0.1), linetype = 2,color = "blue") + 
  geom_line(stat = 'summary', fun.y = quantile, fun.args = list(probs = 0.9), linetype = 2,color = "blue") + 
  geom_line(stat = 'summary', fun.y = quantile, fun.args = list(probs = 0.5), linetype = 1,color = "grey") +
  coord_cartesian(xlim = c(13,90), ylim = c(0,1000)) +
  scale_x_continuous(breaks = seq(10,90,10))
  

```


#### correlation

```{r}
cor(pf$age, pf$friend_count)
cor.test(pf$age, pf$friend_count, method = "pearson")
```

```{r}

with(pf, cor.test(age, friend_count, method = "pearson"))

```

[Correlation (Pearson, Kendall, Spearman)](http://www.statisticssolutions.com/correlation-pearson-kendall-spearman/)

- **Pearson r correlation**, both variables should be normally distributed
- Other assumptions include linearity and homoscedasticity.  Linearity assumes a straight line relationship between each of the variables in the analysis and homoscedasticity assumes that data is normally distributed about the regression line


- **Kendall rank correlation**: Kendall rank correlation is a non-parametric test that measures the strength of dependence between two variables.  If we consider two samples, a and b, where each sample size is n, we know that the total number of pairings with a b is n(n-1)/2.  The following formula is used to calculate the value of Kendall rank correlation:


- **Spearman rank correlation**: Spearman rank correlation is a non-parametric test that is used to measure the degree of association between two variables.  It was developed by Spearman, thus it is called the Spearman rank correlation.  Spearman rank correlation test does not assume any assumptions about the distribution of the data and is the appropriate correlation analysis when the variables are measured on a scale that is at least ordinal.
The following formula is used to calculate the Spearman rank correlation:
spearman rank correlation
P= Spearman rank correlation
di= the difference between the ranks of corresponding values Xi and Yi
n= number of value in each data set


```{r}
ggplot(pf, aes(y = likes_received, x = www_likes_received)) +
  geom_point(color = "orange", alpha = 0.05, position = position_jitter(h = 0)) +
  xlim(0, quantile(pf$www_likes_received, 0.95)) +
  ylim(0, quantile(pf$likes_received, 0.95)) +
  geom_smooth(method = 'lm', color = 'red')

```



```{r}
library(alr3)
data("Mitchell")

ggplot(Mitchell, aes(Temp, Month)) +
  geom_point()


```



```{r}
ggplot(aes(x=(Month%%12),y=Temp),data=Mitchell)+ 
  geom_point() 

```

```{r}
pf <- read.delim('pseudo_facebook.tsv')
pf$age_with_months <- with(pf, age + (1 - dob_month / 12))

pf.fc_by_age_months <- pf %>% 
  group_by(age_with_months) %>% 
  summarise(friend_count_mean = mean(friend_count), friend_count_median = median(friend_count), n = n()) %>% 
  arrange(age_with_months)

pf.fc_by_age_months

ggplot(subset(pf.fc_by_age_months, age_with_months < 71), aes(age_with_months, friend_count_mean)) +
  geom_point()

```
